---
title: "Simulation setup"
output: html_document
date: "2025-11-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Simulation setup

This document runs through the simulation setup to generate the FBD trees and the character matrices

### Importing constants

First we import all the constants needed for the simulation setup

```{r}
source("code/constants.R")
```

## Simulate the tree

The standard setup uses an origination rate of `r lambda`, a extinction rate of `r mu` and a duration of `r t_max` Myr. We simulate the tree, rejecting simulations that do not make it to the present:

```{r}
set.seed(578477)
    repeat{
      tree_complete = TreeSim::sim.bd.age(age = t_max,
                                          numbsim = 1,
                                          lambda = lambda,
                                          mu = mu,
                                          mrca =FALSE,
                                          complete = TRUE)[[1]]
      if (class(tree_complete) == "phylo"){ break }
    }
plot(tree_complete)
```

For curiosity, we can see how often the simulation fails to produce "proper" trees:

```{r}
n = 200
 tree_list = TreeSim::sim.bd.age(age = t_max,
                                          numbsim = n,
                                          lambda = lambda,
                                          mu = mu,
                                          mrca =FALSE,
                                          complete = TRUE)
sapply(tree_list, function(x) class(x) == "phylo") |> sum()/n
# plot histogram of no of tips
sapply(tree_list[sapply(tree_list, function(x) class(x) == "phylo")], Ntip) |> hist(main = "No of tips")




```
For thes selected parameters, approx. 30 % of trees are accepted (depending on the seed)

## Simulate fossils

Next, we simulate the fossils along the tree. We need some utility functions:

```{r}
subsample_fossils = function(fossils, n){
  #' randomly select a number of fossils
  ind = sample(seq_len(length(fossils$sp)), size = n)
  return(fossils[ind,])
}

get_collection_prob = function(adm){
  #' @title determine collection probability
  #' 
  #' @param adm an age-depth modek
  #' 
  #' @description
  #' given an age-depth model, returns a function that returns 0 if the time coincides with a hiatus
  #' and 1 if not. Outside of the domain of the age-depth model, it returns 1 
  #' 
  collection_prob = function(x){
    admtools::is_destructive(adm, x, out_dom_mode = "conservative") |> (\(x){!x})() |> as.numeric()
  }
  return(collection_prob)
}
```


We distinguish between the different different stratigraphic cases:

```{r}
case = "sinusoid"
  source("code/strat_scenarios.R") # load age-depth models stored
    if(case == "sinusoid"){adm = adm_sinusoid}
    if(case == "miller"){adm = adm_miller}
    if (case == "base"){adm = admtools::tp_to_adm(t = c(0,t_max), h = c(0, t_max))}
plot(adm, lty_destr = 0)

```

From the age-depth model, we create a function that specifies the preservation probability of fossils as a function of time:

```{r}
f = get_collection_prob(adm)
t = seq(0, t_max, length.out = 10000)
plot(t, f(t), type = "l", xlab = "Time", ylab = "Preservation prob")
```

We can combine this function with `StratPal::apply_taphonony` to select fossils that don't coincide with gaps:

```{r}
    fossils_full_base = FossilSim::sim.fossils.poisson(rate = sampling_rate,
                                                       tree = tree_complete) |>
      admtools::rev_dir(ref = t_max) |> 
      StratPal::apply_taphonomy(pres_potential = identity, ctc =  f) |>
      admtools::rev_dir(ref = t_max) |>
      subsample_fossils(n = n_fossils)

hist(fossils_full_base$hmin, breaks = seq(0, t_max, by = 0.1), main = "fossil ages", xlab = "Time")
```

Note here that we simulate a lot of fossils, then remove those that coincide with a hiatus, and then select `r n_fossils` of the preserved ones as our fossil record.

## Add fossils to tree

Next the fossils are added to the tree, and determine which tips are extant

```{r}
# add to tree, to assign tip label names to the fossils
    tree_sa_complete = FossilSim::SAtree.from.fossils(tree = tree_complete,
                                                      fossils = fossils_full_base)
    # fossils with tip label names
    fossils_full = tree_sa_complete$fossils
    
    tree_fp_rho1 = FossilSim::sampled.tree.from.combined(tree = tree_sa_complete$tree,
                                                         rho = 1)
    tree_fp_rho0 = FossilSim::sampled.tree.from.combined(tree = tree_sa_complete$tree,
                                                         rho = 0)
    extant_tips = setdiff(tree_fp_rho1$tip.label, tree_fp_rho0$tip.label )
    print(extant_tips)
    
    plot(tree_fp_rho1)
    
```

## simulate morphological and molecular partition

Next we simulate the morphological and molecular partitions using some utility functions:

```{r}
sim_mol_char = function(tree, extant_tips, opts, clock_rate){
  #' @title simulate molecular partition for extant tips
  #' 
  tree_w_rate_mol = tree
  tree_w_rate_mol$edge.length = tree_w_rate_mol$edge.length * clock_rate
  tree_w_rate_mol$root.edge = tree_w_rate_mol$root.edge * clock_rate
  a = phyclust::seqgen(opts = opts, rooted.tree = tree_w_rate_mol) |>
    strsplit(" +")
  l = list()
  for (i in 1:length(a)){
    if (a[[i]][1] %in% extant_tips){
      l[[ a[[i]][1] ]] = unlist(strsplit(a[[i]][2], split = ""))
    }
  }
  return(l)
}

sim_bin_char = function(tree, nchar, clock_rate){
  #' @title simulate discrete traits along tree
  #' 
  #' @param tree tree
  #' @param rate rate for Q matrix
  #' @param nchar number of characters to simulate
  #' 
  tree_w_rate_morph = tree
  tree_w_rate_morph$edge.length = tree_w_rate_morph$edge.length * clock_rate
  tree_w_rate_morph$root.edge = tree_w_rate_morph$root.edge * clock_rate
  par = matrix(data = c(-1, 1, 1, -1), 
               nrow = 2, 
               ncol = 2)
  char_mat = geiger::sim.char(phy = tree_w_rate_morph,
                              par = par,
                              model = "discrete",
                              nsim = nchar)
  char_mat = apply(char_mat, 1, function(x) x -1) |> t()
  return(char_mat)
}
```

For the molecular partition, `opts` specifies the model, which is give by the HKY model, 5 categories for the gamma distribution. alpha shape parameter set to 0.25, transition to transversion ratio is 5, and an alignment of length 2000

For the morphological partition, we use the Mk model (without ascertainment corrections/discarding of non-informative characters), with a varying number of characters

In both cases we assume a strict clock

```{r}
n_char = 30
char_mat = sim_bin_char(tree = tree_fp_rho1,
                            nchar = n_char,
                            clock_rate = clock_rate_morph)

mol = sim_mol_char(tree = tree_fp_rho1,
                       extant_tips = extant_tips,
                       opts = opts,
                       clock_rate = clock_rate_mol)
```

## Writing the data

Now all data can be written out to be inspected. We use a utility function that adds extant tips as fossils with age 0 for usage in revBayes:

```{r}
get_fossil_ages = function(fossils, extant_taxa = NULL){
  #' @title get ages of samples from fossils object
  #' 
  #' @param fossils a fossil object
  #' @param recent_taxa list of taxon names to be added as recent (i.e., with age 0)
  if (is.null(extant_taxa)){
    df = data.frame(taxon = fossils$tip.label,
                    min_age = fossils$hmin,
                    max_age = fossils$hmax)
  }
  if (!is.null(extant_taxa)){
    n_recent = length(extant_taxa)
    df = data.frame(taxon = c(fossils$tip.label, extant_taxa),
                    min_age = c(fossils$hmin, rep(0, n_recent)),
                    max_age = c(fossils$hmax, rep(0, n_recent)))
  }    
  return(df)
}
```


```{r}
    ape::write.nexus.data(mol,
                          file = "mol_dat.nex")
    ape::write.nexus.data(char_mat,
                          file = "morph_data.nex",
                          format = "standard")
    truth = ape::write.nexus(tree_fp_rho1,
                     file= "true_tree.nex")
    
      fossils_full |>
      get_fossil_ages(extant_taxa = extant_tips)|>
      write.table(file = "fossils.csv",
                  quote = FALSE,
                  row.names = FALSE)
    
```


